\chapter*{Acknowledgements}

First of all, I would like to thank my advisor Helmut Veith for his guidance and support during my master's studies. He suggested the topic of this thesis, pointed out relevant research, and provided direction and insight. Most importantly, he recognized and believed in my strengths and gave me opportunity and room to develop – thank you!

Next, I would like to thank my co-advisor Florian Zuleger for his advice and interest in my work. His sense for focusing on the important problems, as well as communicating them clearly in a big-picture view have greatly helped me understand my work in context and relevance to related work. He also put a tremendous amount of time into giving detailed feedback on both content and structure, without which the work at hand would not exist.

This thesis has connections to Yulia Demyanova's work on variable roles, and Moritz Sinn's work on bound computation. Thank you both for discussing your work with me, and for sharing your experience with implementing Clang- and LLVM-based tools. In general, I'm grateful to all my colleagues from the FORSYTE group and on the "third floor" for this pleasant work environment.

Pursuing an academic degree \emph{and} various other interests, one needs a stable frame to support the picture. I would like to thank my parents and grandparents for their continued support of my plans. My sisters Carina and Christina have held the frame as well as painted the picture with me. Thank you for your love, understanding, and the possibility to discuss just about anything.

Along the way, many more people have considered it worth their time to teach me. Even though their number is finite, I cannot possibly hope to compile an exhaustive list. I am greatly indebted to all of you, for sharing your wisdom and for shaping my life.

\vspace{3em}

\begin{flushright}
Vienna, December 2013\\
Thomas Pani
\end{flushright}
